---
title: "The Theoretical Minimum of Classical Mechanics Notebook"
author: "Geirmundur Orri Sigurdsson"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is my notebook for studying of The Theoretical Minimum of Classical Mechanics by Leonard Susskind and George Hrabovsky (2013) ISBN 978-0-465-07568-3

# Dependencies

```{r dependencies, message=FALSE, warning=FALSE}
library(plotly)
library(ggplot2)
```

# Interlude 1: Spaces, Trigonometry, and Vectors

## Mathematics

### Trigonometry

$$\texttt{sin}\left ( \alpha + \beta \right ) = \texttt{sin} \alpha \texttt{cos} \beta + \texttt{cos} \alpha \texttt{sin} \beta$$
$$\texttt{sin}\left ( \alpha - \beta \right ) = \texttt{sin} \alpha \texttt{cos} \beta - \texttt{cos} \alpha \texttt{sin} \beta$$
$$\texttt{cos}\left ( \alpha + \beta \right ) = \texttt{cos} \alpha \texttt{cos} \beta - \texttt{sin} \alpha \texttt{sin} \beta$$
$$\texttt{cos}\left ( \alpha - \beta \right ) = \texttt{cos} \alpha \texttt{cos} \beta + \texttt{sin} \alpha \texttt{sin} \beta$$
$$\texttt{sin}^{2} \theta + \texttt{cos}^{2} \theta = 1$$

### Vectors

$$\vec{V} = V_{x}\hat{i} + V_{y}\hat{j} + V_{z}\hat{k}$$
$$\left |\vec{V} \right | = \sqrt{ V_{x}^{2} + V_{y}^{2} + V_{z}^{2} }$$
$$a\vec{V} = \left ( aV_{x}, aV_{y}, aV_{z} \right )$$
$$\left ( \vec{A} + \vec{B} \right )_{x} = \left ( A_{x} + B_{x} \right )$$
$$\left ( \vec{A} + \vec{B} \right )_{y} = \left ( A_{y} + B_{y} \right )$$
$$\left ( \vec{A} + \vec{B} \right )_{z} = \left ( A_{z} + B_{z} \right )$$
$$\vec{A} \cdot \vec{B} = \left |\vec{A} \right | \left |\vec{B} \right | \texttt{cos} \theta$$
$$\vec{A} \cdot \vec{B} = A_{x}B_{x} + A_{y}B_{y} + A_{z}B_{z}$$
If $$\vec{A} \cdot \vec{B} = 0$$ then $$\vec{A}$$ and $$\vec{B}$$ are orthogonal

## Exercise 1

```{r interlude 1 exercise 1}
f <- function(t) {
  t * t * t * t + 3 * t * t * t - 12 * t * t - 6
}
t <- seq(from = -4, to = 4, by = 0.01)
y <- f(t)
plot(t, y)

g <- function(x) {
  sin(x) - cos(x)
}
x <- seq(from = 0, to = 2 * pi, by = 0.01)
y <- g(x)
plot(x, y)

t <- function(a) {
  exp(a) + a * log(a)
}
a <- seq(from = 0, to = 5, by = 0.01)
y <- t(a)
plot(a, y)

x <- function(t) {
  sin(t) * sin(t) - cos(t)
}
t <- seq(from = 0, to = 2 * pi, by = 0.01)
y <- x(t)
plot(t, y)
```

## Exercise 4

```{r interlude 1 exercise 4}
Ax <- 2
Ay <- -3
Az <- 1
Bx <- -4
By <- -3
Bz <- 2
Am <- sqrt(Ax * Ax + Ay * Ay + Az * Az)
Bm <- sqrt(Bx * Bx + By * By + Bz * Bz)
Dp <- Ax * Bx + Ay * By + Az * Bz
radians <- acos(Dp/(Am * Bm))
degrees <- radians * 180 / pi
degrees
```

## Exercise 5

```{r interlude 1 exercise 5}
dot_product <- function(a, b) {
  a[1] * b[1] + a[2] * b[2] + a[3] * b[3]
}
A <- c( 1,  1, 1)
B <- c( 2, -1, 3)
C <- c( 3,  1, 0)
D <- c(-3,  0, 2)
dot_product(A, B)
dot_product(A, C)
dot_product(A, D)
dot_product(B, C)
dot_product(B, D) # B and D are orthogonal because the dot product is zero
dot_product(C, D)
```

# Lecture 2: Motion

## Mathematics

$$\frac{\mathrm{d}f\left ( t \right )}{\mathrm{d} t} = f{}'\left ( t \right ) = \lim_{\Delta t \to 0}\frac{\Delta f}{\Delta t} = \lim_{\Delta t \to 0}\frac{f\left ( t + \Delta t \right ) - f\left ( t \right )}{\Delta t}$$
$$\frac{\mathrm{d}\left ( t^{n} \right )}{\mathrm{d} t} = nt^{n - 1}$$
$$\frac{\mathrm{d}c}{\mathrm{d} t} = 0$$
$$\frac{\mathrm{d}\left (cf \right )}{\mathrm{d} t} = c \frac{\mathrm{d}f}{\mathrm{d} t}$$
$$\frac{\mathrm{d}\left (\texttt{sin} t \right )}{\mathrm{d} t} = \texttt{cos} t$$
$$\frac{\mathrm{d}\left (\texttt{cos} t \right )}{\mathrm{d} t} = -\texttt{sin} t$$
$$\frac{\mathrm{d}\left (\texttt{e}^{t} \right )}{\mathrm{d} t} = \texttt{e}^{t}$$
$$\frac{\mathrm{d}\left (\texttt{ln} t \right )}{\mathrm{d} t} = \frac{1}{t} $$
Sum rule
$$\frac{\mathrm{d}\left ( f + g \right )}{\mathrm{d} t} = \frac{\mathrm{d}\left ( f \right )}{\mathrm{d} t} + \frac{\mathrm{d}\left ( g \right )}{\mathrm{d} t}$$
Product rule
$$\frac{\mathrm{d}\left ( f g \right )}{\mathrm{d} t} = f \left ( t \right ) \frac{\mathrm{d} g }{\mathrm{d} t} + g\left ( t \right ) \frac{\mathrm{d}f }{\mathrm{d} t}$$
Chain rule
$$\frac{\mathrm{d} f}{\mathrm{d} t} = \frac{\mathrm{d} f }{\mathrm{d} g} \frac{\mathrm{d} g }{\mathrm{d} t}$$
Some frequently used derivatives
$$\frac{\mathrm{d}\left (\texttt{sin}^{2} t \right )}{\mathrm{d} t} = 2 \texttt{sin} t \texttt{cos} t$$
$$\frac{\mathrm{d}\left (\texttt{cos}^{2} t \right )}{\mathrm{d} t} = -2 \texttt{sin} t \texttt{cos} t$$
$$\frac{\mathrm{d}\left (\texttt{sin}^{3} t \right )}{\mathrm{d} t} = 3 \texttt{sin}^{2} t \texttt{cos} t$$
$$\frac{\mathrm{d}\left (\texttt{cos}^{3} t \right )}{\mathrm{d} t} = -3 \texttt{cos}^{2} t \texttt{sin} t$$

## Exercise 1 and 2

The first and second derivatives of each of the following functions.

$$f\left ( t \right ) = t^{4} + 3t^{3} - 12t^{2} + t - 6$$
$$\frac{\mathrm{d}f}{\mathrm{d} t} = f{}'\left ( t \right ) = 4t^{3} + 9t^{2} - 24t + 1$$
$$\frac{\mathrm{d}^{2}f}{\mathrm{d} t^{2}} = f{}'{}'\left ( t \right ) = 12t^{2} + 18t - 24$$

$$g\left ( x \right ) = \texttt{sin}x - \texttt{cos}x$$
$$\frac{\mathrm{d}g}{\mathrm{d} x} = g{}'\left ( x \right ) = \texttt{cos}x + \texttt{sin}x$$
$$\frac{\mathrm{d}^{2}g}{\mathrm{d} x^{2}} = g{}'{}'\left ( x \right ) = -\texttt{sin}x + \texttt{cos}x = \texttt{cos}x - \texttt{sin}x = -g\left ( x \right )$$

$$\theta\left ( \alpha \right ) = \texttt{e}^{\alpha} + \alpha \texttt{ln}\alpha$$
$$\frac{\mathrm{d}\theta}{\mathrm{d} \alpha} = \theta{}'\left ( \alpha \right ) = \texttt{e}^{\alpha} + \alpha \frac{1}{\alpha} + \texttt{ln}\alpha = \texttt{e}^{\alpha} + \texttt{ln}\alpha + 1$$
$$\frac{\mathrm{d}^{2}\theta}{\mathrm{d} \alpha^{2}} = \theta{}'{}'\left ( \alpha \right ) = \texttt{e}^{\alpha} + \frac{1}{\alpha}$$

$$x\left ( t \right ) = \texttt{sin}^{2}t - \texttt{cos}t = \texttt{sin}t \cdot  \texttt{sin}t - \texttt{cos}t$$
$$\frac{\mathrm{d}x}{\mathrm{d} t} = x{}'\left ( t \right ) = \texttt{sin}t \cdot \texttt{cos}t + \texttt{sin}t \cdot \texttt{cos}t + \texttt{sin}t = 2 \texttt{sin}t \texttt{cos}t + \texttt{sin}t$$
$$\frac{\mathrm{d}^{2}x}{\mathrm{d} t^{2}} = x{}'{}'\left ( t \right ) = 2 \left ( \texttt{sin}t \cdot \left ( -\texttt{sin}t \right ) + \texttt{cos}t \cdot \texttt{cos}t \right ) + \texttt{cos}t = 2\texttt{cos}^{2}t - 2\texttt{sin}^{2}t + \texttt{cos}t$$

## Exercise 8

$$\vec{r} = \left ( \texttt{cos} \omega t, \texttt{e}^{\omega t} \right )$$
```{r lecture 2 exercise 8 ar}
o <- pi / 10
t <- seq(from = -10, to = 10, by = 0.01)

rx <- function(t, o) { cos(o * t) }
ry <- function(t, o) { exp(o * t) }
x <- rx(t, o)
y <- ry(t, o)
plot(t, x)
plot(t, y)
pd <- data.frame(t, x, y)
ggplot(data = pd, mapping = aes(x = x, y = y)) + geom_point(aes(color = t)) + theme_light()
plot_ly(data = pd, type = 'scatter3d', mode = 'markers', x = ~x, y = ~y, z = ~t, color = ~t)
```

$$\vec{v} = \dot{\vec{r}} = \left ( -\omega\texttt{sin} \omega t, \omega \texttt{e}^{\omega t} \right )$$
```{r lecture 2 exercise 8 av}
vx <- function(t, o) { -o * sin(o * t) }
vy <- function(t, o) { o * exp(o * t) }
x <- vx(t, o)
y <- vy(t, o)
plot(t, x)
plot(t, y)
pd <- data.frame(t, x, y)
ggplot(data = pd, mapping = aes(x = x, y = y)) + geom_point(aes(color = t)) + theme_light()
plot_ly(data = pd, type = 'scatter3d', mode = 'markers', x = ~x, y = ~y, z = ~t, color = ~t)
```

$$\vec{a} = \ddot{\vec{r}} = \left ( -\omega^{2} \texttt{cos} \omega t, \omega^{2} \texttt{e}^{\omega t} \right )$$
```{r lecture 2 exercise 8 aa}
ax <- function(t, o) { -o * o * cos(o * t) }
ay <- function(t, o) { o * o * exp(o * t) }
x <- ax(t, o)
y <- ay(t, o)
plot(t, x)
plot(t, y)
pd <- data.frame(t, x, y)
ggplot(data = pd, mapping = aes(x = x, y = y)) + geom_point(aes(color = t)) + theme_light()
plot_ly(data = pd, type = 'scatter3d', mode = 'markers', x = ~x, y = ~y, z = ~t, color = ~t)
```

$$\vec{r} = \left ( \texttt{cos} \left ( \omega t - \phi \right ), \texttt{sin} \left ( \omega t - \phi \right ) \right )$$
```{r lecture 2 exercise 8 br}
omega <- pi / 10
phi <- pi / 10
t <- seq(from = -10, to = 10, by = 0.01)

rx <- function(t, omega, phi) { cos(omega * t - phi) }
ry <- function(t, omega, phi) { sin(omega * t - phi) }
x <- rx(t, omega, phi)
y <- ry(t, omega, phi)
plot(t, x)
plot(t, y)
pd <- data.frame(t, x, y)
ggplot(data = pd, mapping = aes(x = x, y = y)) + geom_point(aes(color = t)) + theme_light()
plot_ly(data = pd, type = 'scatter3d', mode = 'markers', x = ~x, y = ~y, z = ~t, color = ~t)
```

$$\vec{v} = \dot{\vec{r}} = \left ( -\omega \texttt{sin} \left ( \omega t - \phi \right ), \omega \texttt{cos} \left ( \omega t - \phi \right ) \right )$$
```{r lecture 2 exercise 8 bv}
vx <- function(t, omega, phi) { -omega * sin(omega * t - phi) }
vy <- function(t, omega, phi) { omega * cos(omega * t - phi) }
x <- vx(t, omega, phi)
y <- vy(t, omega, phi)
plot(t, x)
plot(t, y)
pd <- data.frame(t, x, y)
ggplot(data = pd, mapping = aes(x = x, y = y)) + geom_point(aes(color = t)) + theme_light()
plot_ly(data = pd, type = 'scatter3d', mode = 'markers', x = ~x, y = ~y, z = ~t, color = ~t)
```

$$\vec{a} = \ddot{\vec{r}} = \left ( -\omega^{2} \texttt{cos} \left ( \omega t - \phi \right ), -\omega^{2} \texttt{sin} \left ( \omega t - \phi \right ) \right )$$
```{r lecture 2 exercise 8 ba}
ax <- function(t, omega, phi) { -omega * omega * cos(omega * t - phi) }
ay <- function(t, omega, phi) { -omega * omega * sin(omega * t - phi) }
x <- ax(t, omega, phi)
y <- ay(t, omega, phi)
plot(t, x)
plot(t, y)
pd <- data.frame(t, x, y)
ggplot(data = pd, mapping = aes(x = x, y = y)) + geom_point(aes(color = t)) + theme_light()
plot_ly(data = pd, type = 'scatter3d', mode = 'markers', x = ~x, y = ~y, z = ~t, color = ~t)
```

$$\vec{r} = \left ( c\texttt{cos}^{3}t, c\texttt{sin}^{3}t \right )$$
```{r lecture 2 exercise 8 cr}
c <- 0.01
t <- seq(from = -10, to = 10, by = 0.01)

rx <- function(t, c) { c * cos(t) * cos(t) * cos(t) }
ry <- function(t, c) { c * sin(t) * sin(t) * sin(t) }
x <- rx(t, c)
y <- ry(t, c)
plot(t, x)
plot(t, y)
pd <- data.frame(t, x, y)
ggplot(data = pd, mapping = aes(x = x, y = y)) + geom_point(aes(color = t)) + theme_light()
plot_ly(data = pd, type = 'scatter3d', mode = 'markers', x = ~x, y = ~y, z = ~t, color = ~t)
```

$$\vec{v} = \dot{\vec{r}} = \left ( -3c\texttt{cos}^{2}t\texttt{sin}t, 3c\texttt{sin}^{2}t\texttt{cos}t \right )$$
```{r lecture 2 exercise 8 cv}
vx <- function(t, c) { -3 * c * cos(t) * cos(t) * sin(t) }
vy <- function(t, c) {  3 * c * sin(t) * sin(t) * cos(t) }
x <- vx(t, c)
y <- vy(t, c)
plot(t, x)
plot(t, y)
pd <- data.frame(t, x, y)
ggplot(data = pd, mapping = aes(x = x, y = y)) + geom_point(aes(color = t)) + theme_light()
plot_ly(data = pd, type = 'scatter3d', mode = 'markers', x = ~x, y = ~y, z = ~t, color = ~t)
```

$$\vec{a} = \ddot{\vec{r}} = \left ( 6c\texttt{sin}^{2}t\texttt{cos}t - 3c\texttt{cos}^{3}t, 6c\texttt{cos}^{2}t\texttt{sin}t - 3c\texttt{sin}^{3}t \right )$$
```{r lecture 2 exercise 8 ca}
ax <- function(t, c) { 6 * c * sin(t) * sin(t) * cos(t) - 3 * c * cos(t) * cos(t) * cos(t) }
ay <- function(t, c) { 6 * c * cos(t) * cos(t) * sin(t) - 3 * c * sin(t) * sin(t) * sin(t) }
x <- ax(t, c)
y <- ay(t, c)
plot(t, x)
plot(t, y)
pd <- data.frame(t, x, y)
ggplot(data = pd, mapping = aes(x = x, y = y)) + geom_point(aes(color = t)) + theme_light()
plot_ly(data = pd, type = 'scatter3d', mode = 'markers', x = ~x, y = ~y, z = ~t, color = ~t)
```

$$\vec{r} = \left ( c \left ( t - \texttt{sin}t \right ), c \left ( 1 - \texttt{cos}t \right ) \right )$$
```{r lecture 2 exercise 8 dr}
c <- 0.01
t <- seq(from = -10, to = 10, by = 0.01)

rx <- function(t, c) { c * ( t - sin(t)) }
ry <- function(t, c) { c * ( 1 - cos(t)) }
x <- rx(t, c)
y <- ry(t, c)
plot(t, x)
plot(t, y)
pd <- data.frame(t, x, y)
ggplot(data = pd, mapping = aes(x = x, y = y)) + geom_point(aes(color = t)) + theme_light()
plot_ly(data = pd, type = 'scatter3d', mode = 'markers', x = ~x, y = ~y, z = ~t, color = ~t)
```

$$\vec{v} = \dot{\vec{r}} = \left ( c \left ( 1 - \texttt{cos}t \right ), c \left ( \texttt{sin}t \right ) \right )$$
```{r lecture 2 exercise 8 dv}
vx <- function(t, c) { c * ( 1 - cos(t)) }
vy <- function(t, c) { c * ( sin(t)) }
x <- vx(t, c)
y <- vy(t, c)
plot(t, x)
plot(t, y)
pd <- data.frame(t, x, y)
ggplot(data = pd, mapping = aes(x = x, y = y)) + geom_point(aes(color = t)) + theme_light()
plot_ly(data = pd, type = 'scatter3d', mode = 'markers', x = ~x, y = ~y, z = ~t, color = ~t)
```

$$\vec{a} = \ddot{\vec{r}} = \left ( c \left ( \texttt{sin}t \right ), c \left ( \texttt{cos}t \right ) \right )$$
```{r lecture 2 exercise 8 da}
ax <- function(t, c) { c * ( sin(t)) }
ay <- function(t, c) { c * ( cos(t)) }
x <- ax(t, c)
y <- ay(t, c)
plot(t, x)
plot(t, y)
pd <- data.frame(t, x, y)
ggplot(data = pd, mapping = aes(x = x, y = y)) + geom_point(aes(color = t)) + theme_light()
plot_ly(data = pd, type = 'scatter3d', mode = 'markers', x = ~x, y = ~y, z = ~t, color = ~t)
```